Phenotype and risk factors of venom-induced anaphylaxis: a case-control study of the European Anaphylaxis Registry

### 2020-03-27 analysis/paper/paper.Rmd
small correction for JACI

---
title: "Phenotype and risk factors of venom-induced anaphylaxis: a case-control study of the European Anaphylaxis Registry"
output: 
    #bookdown::html_document2:
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/vancouver.csl" # Insert path for the bib-style
---

Wojciech Francuzik^1^,
Franziska Ruëff^2^,
Andrea Bauer^3^,.
Maria Beatrice Bilò ^4,5^,
Victoria Cardona^6^,
George Christoff^7^,
Sabine Dölle-Bierke^8^,
Luis Ensina^9^,
Montserat Fernandes-Rivas^7,11^,
Thomas Hawranek^12^,
Johnathan O'B Hourihane^13^,
Tilo Jakob^14,15^,
Nicos G. Papadopoulus^16,17^,
Claudia Pföhler^18^,
Iwona Poziomkowska-Gęsicka^19^,
Xavier Van der Brempt^20^,
Kathrin Scherer Hofmeier^21^,
Regina Treudler^22^,
Nicola Wagner^23^,
Bettina Wedi^24^,
Margitta Worm^1^

**Affiliations:**

@@

**Corresponding author: **
  Prof. Dr. med. M. Worm margitta.worm@charite.de
Phone: +49 30 450 529 005;      Fax: +49 30 450 529 902

<!-- **Highlights:** -->
<!--   These are the highlights.  -->
**Document statistics:**
2838 words, 5 figures, 0 tables, 29 references

@@

# Key points

**Question:** What is the phenotype of insect venom anaphylaxis?

**Findings:** This case-control study showed that venom-induced anaphylaxis significantly more often presented with cardiovascular symptoms. Secondly, severe cases of venom-induced anaphylaxis were significantly more often associated with the lack of skin symptoms upon anaphylaxis and showed a significant interaction with the levels of baseline serum tryptase (in range from 8 - 11 ng/ml).

**Meaning:** Allegologists should consider intesified prophylaxis (and diagnostic) measures in patients with baseline tryptase levels of above 8 and a history of insect venom anaphylaxis.

@@

**Importance:** Venom-induced anaphylaxis is a common, potentially life-threatening hypersensitivity reaction associated with specific: 1) symptom profile, 2) cofactors, and 3) management. Identifying the differences in phenotypes of anaphylaxis is crucial for future management guidelines and the development of a personalized medicine approach.

**Objective:** This study aimed to evaluate the phenotype and risk factors of venom-induced anaphylaxis in comparison to other elicitors of anaphylaxis.

**Design:** Using data from the European Anaphylaxis Registry (`r nrow(data)` cases) we identified `r length(which(data4$d_elicitor_gr5== "insects" & data4$reaction_type_brown=="anaphylaxis"))` with venom-induced anaphylaxis and analyzed these in comparison to sex- and age-matched anaphylaxis cases triggered by other elicitors (non-VIA n = `r nrow(age_sex_matched %>% filter(grouping == "other"))`).

**Setting:** Multicenter, international, prospective, case-control study.

**Participants:** All patients with the diagnosis of moderate to severe anaphylaxis (according to Ring and Messmer grade II-IV) who gave informed consent to provide their clinical data for the Anaphylaxis Registry were included in the study. Subsequently, we matched the group of patients presenting with venom anaphylaxis to sex- and age-matched group of patients who presented with anaphylaxis to other elicitors (i.e., food, drugs).

**Main Outcome(s) and Measure(s):** This study measured the severity of anaphylaxis on the Ring and Messmer scale. Hypotheses regarding the comparisons between groups were formulated during and after data collection.

@@

**Conclusions and Relevance:** Regarding the fact that venom-induced anaphylaxis frequently affects the cardiovascular system, patients were undertreated with epinephrine. The lack of skin symptoms (i.e., urticaria or flushing) during anaphylaxis and baseline serum tryptase levels within the upper normal limits were associated with severe reactions. Patients with serum tryptase of above 8 ng/ml and a history of venom anaphylaxis may require additional prophylaxis.

@@

BST levels were significantly higher in patients with a prior diagnosis of mastocytosis \@ref(fig:BST). 
Next, we investigated the association of skin symptoms with the tryptase levels in non-mastocytosis patients. For this model, we excluded the cases with known mastocytosis and with BST above 11.5 ng/ml, potentially indicating non-diagnosed mast cell activation disorders. Similarly, 1) tryptase levels were higher in VIA patients, 2) correlated with the severity of anaphylaxis, and 3) this effect was significant in VIA (`r skin_symptoms$fit_tryp[[1]][4,5] %>% pval()`) but not in the non-VIA group (Fig. \@ref(fig:figskinsymptoms)C-D).

@@

The cofactor most prominently associated with an increased risk of severe anaphylaxis was mastocytosis (Fig. \@ref(fig:figcofactors)). Concomitant mastocytosis increased the risk for 1) cardiac arrest and 2) loss of consciousness in patients undergoing VIA significantly more than in patients undergoing anaphylaxis due to other elicitors (Fig. \@ref(fig:figcofactors)C and Fig. S\@ref(fig:supplement)A).

@@

Concomitant cardiovascular diseases were more prevalent in VIA than in non-VIA cases (`r age_sex_matched %>% count(grouping,q_410_cardio_cur) %>% group_by(grouping) %>% mutate(prop= n/sum(n)) %>% {.[c(2,5),4]} %>% pull() %>% compare(x = (age_sex_matched %>% count(grouping,q_410_cardio_cur) %>% group_by(grouping) %>% {.[c(2,5),3]} %>% pull()), type = "both")`) and were associated with higher risk of severe anaphylaxis when elicited by insects but were not relevant in non-VIA cases (Fig. \@ref(fig:figcofactors)). Interestingly, BST values were increased in patients with concomitant cardiovascular diseases, irrespectively of the reaction severity (Fig. S\@ref(fig:trypcardiocur)).

@@

Severe reactions of VIA were more prevalent in patients above 22 years of age, and in VIA cases vs. non-VIA cases (Fig. S\@ref(fig:sev)). There were no differences in severity of reactions elicited by yellow-jackets and other 
insect species (p = 0.4128).

@@

IVA was more often severe if the reaction occured in the first 10 minutes after exposure to venom (`r sev_time$insects[1,2] %>% {.*100} %>% round(2)`% of Ring and Messmer grade III or IV cases) then when the reaction occured after 10 minutes post exposure (`r sev_time$insects[2,2] %>% {.*100} %>% round(2)`% p = `r sev_time$insects_p`).

@@

A positive history of anaphylaxis influenced the therapy of a current episode as well. Epinephrine as a first-line treatment was given less often in VIA cases when compared to other cases **if patients did not report a previous history of anaphylaxis** (`r pvaladrenprev_no`), but in patients reporting previous reactions, there was no difference in epinephrine therapy (`r pvaladrenprev`, Fig. \@ref(fig:adrenalineuse)B). 
Similarly, there were no differences in the epinephrine use between VIA and non-VIA when only severe reactions were taken into consideration (`r adren_severity_sev`). However, when we restricted the analysis to moderate anaphylaxis cases --- non-VIA patients received epinephrine more frequently than VIA (`r adren_severity_mild`).  The presence of skin symptoms during these mild reactions also was associated with a lower fraction of epinephrine treated patients (Fig. \@ref(fig:rfplot)).

Patients with VIA received corticosteroids and antihistamines significantly more frequently than patients with anaphylaxis to other elicitors. On the other hand, epinephrine, beta-2 mimetics, and oxygen were given more often to patients suffering from non-VIA  (Fig. \@ref(fig:adrenalineuse)A, Fig. S\@ref(fig:rfplot)).

@@

Age is an important risk factor for severe anaphylaxis [@Stoevesandt2019]. Adult patients experienced VIA more frequently. Young patients mainly suffer from food-induced anaphylaxis [@Grabenhenrich2016]. 
Emergency room (ER) admission data indicate that the frequency of insect stings hypersensitivity reactions in children is comparable to food hypersensitivity reactions (12-15% of cases of hypersensitivity reactions admitted to the ER), but pediatric anaphylaxis is triggered significantly more often by food elicitors (56% of food hypersensitivity cases vs. 5.3% of sting cases seen in the ER) [@Braganza2005]. Senior patients, on the other hand, suffer from drug-related hypersensitivity more often than insect sting hypersensitivity [@Aurich2019]. <!-- Risk factors - Concomitant disease -->
Similarly, we observed less VIA in patients with concomitant atopic diseases (Fig. \@ref(fig:supplement)) , as these patients more often present with food anaphylaxis [@Tham2019].

@@

We thank all patients, parents, and their children for their support in providing data on the occurrence of anaphylaxis for this study. We thank the study personnel for patients counseling and data entry, and we thank the members of The European Anaphylaxis Registry in detail:

@@

```{r,include=F}
tiff("figS1.tiff", width = 7, height = 4.5,units = 'in', res = 300)
fig_basic
dev.off()
```

@@

```{r,include=F}
tiff("figS2.tiff", width = 7, height = 5,units = 'in', res = 300)
matching.result 
dev.off()
```

@@

```{r,include=F}
tiff("figS3.tiff", width = 7, height = 5,units = 'in', res = 300)
figS3
dev.off()
```

@@

```{r,include=F}
tiff("figS4.tiff", width = 7, height = 4,units = 'in', res = 300)
plot_tryptase_in_cardio_cur
dev.off()
```

@@

```{r sev, fig.cap="Severity of anaphylaxis in subgroups. The severity of patients with VIA in two age groups (left), according to elicitor type (center) and according to the responsible insect species (right)"}
figS5 <- severity_joined_brown %>%
    ggbarplot(x = "grouping",
              y = "n",
              fill = "d_severity_rm",
              position = position_fill(reverse = F),
              color = "d_severity_rm",
              palette = c("#848484","#1f1f1f"))+
    facet_grid(.~subset,scale = "free_x",space = "free_x")+
    labs(fill = "Severity")+
    theme(axis.title.x =  element_blank(),
          axis.text.x = element_text(angle=20,hjust = 1),
          legend.position = "right"
    )+
    #annotate("text",x = 1.5, y = 0.5, label = c("*"),
     #        size =12)+
    guides(color = F)
```

```{r,include=F}
tiff("figS5.tiff", width = 7, height = 4,units = 'in', res = 300)
figS5
dev.off()
```

```{r rfplot, fig.cap="Therapy of anaphylaxis. A: Patients who presented with skin symptoms and VIA less often received epinephrine than if skin symptoms were absent during the reaction. B: Variable importance in the unsupervised classification between VIA and non-VIA using Random Forest classifier.", fig.height=3.5, fig.width=7.1}
figS6 <- ggarrange(
  # adren_prev+
  #   geom_segment(aes(x = 0.5,
  #                    xend = 2.5,
  #                    y = 0.33,
  #                    yend = 0.33),
  #                color="white",
  #                linetype =2),
  skin_sympt_adrenaline,
  rf_ggplot_treatment,
  nrow = 1,
  ncol = 2,
  labels=c("A","B"),
  widths = c(1,1.3)
)
```

```{r,include=F}
tiff("figS6.tiff", width = 7., height = 3.5,units = 'in', res = 300)
figS6
dev.off()
```

@@

```{r,include=F}
tiff("figS7.tiff", width = 7, height = 4,units = 'in', res = 300)
BST_masto$plot
dev.off()
```

>>>>

---
title: "Phenotype and risk factors of venom-induced anaphylaxis: a case-control study of the European Anaphylaxis Registry"
output: 
    #bookdown::html_document2:
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/the-journal-of-allergy-and-clinical-immunology.csl" # Insert path for the bib-style
---

Wojciech Francuzik, MD ^1^, Franziska Ruëff, MD ^2^ , Andrea Bauer, MD ^3^ , Maria Beatrice Bilò, MD
^4,5^ , Victoria Cardona, MD, PhD ^6,7^ , George Christoff, MD, PhD ^8,9^ , Sabine Dölle-Bierke, PhD ^1^ , Luis
Ensina, MD, PhD ^10^ , Montserat Fernandes-Rivas, MD, PhD ^7,11^ , Thomas Hawranek, MD ^12^ ,
Johnathan O’B Hourihane, MD, PhD ^13^ , Thilo Jakob, MD ^14,15^ , Nicos G. Papadopoulus, MD
PhD ^16,17^ , Claudia Pföhler MD ^18^ , Iwona Poziomkowska-Gęsicka, MD, PhD ^19^ , Xavier Van der
Brempt, MD ^20^ , Kathrin Scherer Hofmeier, MD ^21^ , Regina Treudler, MD ^226 , Nicola Wagner, MD ^23^ ,

Bettina Wedi, MD ^24^ , Margitta Worm, MD ^1^

@@

**Corresponding author: **
  Prof. Dr. med. M. Worm, margitta.worm@charite.de,
Phone: +49 30 450 529 005;      Fax: +49 30 450 529 902

<!-- **Highlights:** -->
<!--   These are the highlights.  -->
**Document statistics:**
3136 words, 5 figures, 0 tables, 34 references

@@

<!-- # Key points  -->

<!-- **Question:** What is the phenotype of insect venom anaphylaxis? -->

<!-- **Findings:** This case-control study showed that venom-induced anaphylaxis significantly more often presented with cardiovascular symptoms. Secondly, severe cases of venom-induced anaphylaxis were significantly more often associated with the lack of skin symptoms upon anaphylaxis and showed a significant interaction with the levels of baseline serum tryptase (in range from 8 - 11 ng/ml).   -->

<!-- **Meaning:** Allegologists should consider intesified prophylaxis (and diagnostic) measures in patients with baseline tryptase levels of above 8 and a history of insect venom anaphylaxis.  -->

<!-- ##### -->

@@

**Background:** Venom-induced anaphylaxis is a common, potentially life-threatening
hypersensitivity reaction associated with specific: 1) symptom profile, 2) cofactors, and 3)
management. Identifying the differences in phenotypes of anaphylaxis is crucial for future
management guidelines and the development of a personalized medicine approach.

**Objective:** This study aimed to evaluate the phenotype and risk factors of venom-induced
anaphylaxis.

**Methods:** Using data from the European Anaphylaxis Registry (12874 cases) we identified
3612 with venom-induced anaphylaxis and analyzed these in comparison to sex- and age-
matched anaphylaxis cases triggered by other elicitors (non-VIA n = 3605). All patients
with anaphylaxis (Ring and Messmer grade II-IV) who gave informed consent were
included in the study. Subsequently, we matched the group of patients presenting with
venom anaphylaxis to sex- and age-matched group of patients who presented with
anaphylaxis to other elicitors (i.e., food, drugs).

@@

**Conclusion:** Regarding the fact that venom-induced anaphylaxis frequently affects the
cardiovascular system, patients were undertreated with epinephrine. The lack of skin
symptoms (i.e., urticaria or flushing) during anaphylaxis and baseline serum tryptase levels
within the upper normal limits were associated with severe reactions. Patients with serum
tryptase of above 8 ng/ml and a history of venom anaphylaxis may require additional
prophylaxis.

# Clinical Implications

Allergologists should consider intensified prophylactic (and diagnostic) measures in patients with baseline serum tryptase of above eight ng/ml and a history of insect venom anaphylaxis without skin involvement.

# Capsule Summary

Venom-induced anaphylaxis significantly more often presented with cardiovascular symptoms. Severe cases more often showed lack of skin involvement and were associated with higher levels of baseline serum tryptase (in range from 8 - 11 ng/ml).

**List of abbreviations:**
* VIA - venom-induced anaphylaxis
* BST - baseline serum tryptase
* EAI - epinephrine auto-injector
* MCAS - mast cell activation syndrome
* ER - emergency room

@@

BST levels were significantly higher in patients with a prior diagnosis of mastocytosis (eFig. \@ref(fig:BST)).
We investigated the association of skin symptoms with the tryptase levels in non-mastocytosis patients. For this model, we excluded the cases with known mastocytosis and with BST above 11.5 ng/ml, potentially indicating non-diagnosed mast cell activation disorders. Similarly, 1) tryptase levels were higher in VIA patients, 2) correlated with the severity of anaphylaxis, and 3) this effect was significant in VIA (`r skin_symptoms$fit_tryp[[1]][4,5] %>% pval()`) but not in the non-VIA group (Fig. \@ref(fig:figskinsymptoms)C-D).

@@

The cofactor most prominently associated with an increased risk of severe anaphylaxis was mastocytosis (Fig. \@ref(fig:figcofactors)). Concomitant mastocytosis increased the risk for 1) cardiac arrest and 2) loss of consciousness in patients undergoing VIA significantly more than in patients undergoing anaphylaxis due to other elicitors (Fig. \@ref(fig:figcofactors)C and eFig. \@ref(fig:supplement)A).

@@

Concomitant cardiovascular diseases were more prevalent in VIA than in non-VIA cases (`r age_sex_matched %>% count(grouping,q_410_cardio_cur) %>% group_by(grouping) %>% mutate(prop= n/sum(n)) %>% {.[c(2,5),4]} %>% pull() %>% compare(x = (age_sex_matched %>% count(grouping,q_410_cardio_cur) %>% group_by(grouping) %>% {.[c(2,5),3]} %>% pull()), type = "both")`) and were associated with higher risk of severe anaphylaxis when elicited by insects but were not relevant in non-VIA cases (Fig. \@ref(fig:figcofactors)). Interestingly, BST values were increased in patients with concomitant cardiovascular diseases, irrespectively of the reaction severity (eFig. \@ref(fig:trypcardiocur)).

@@

Severe reactions of VIA were more prevalent in patients above 22 years of age, and in VIA cases vs. non-VIA cases (eFig. \@ref(fig:sev)). There were no differences in severity of reactions elicited by yellow-jackets and other 
insect species (p = 0.4128).

@@

IVA was more often severe if the reaction occured in the first 10 minutes after exposure to venom (`r sev_time$insects[1,2] %>% {.*100} %>% round(2)`% were severe cases) then when the reaction occured after 10 minutes post exposure (`r sev_time$insects[2,2] %>% {.*100} %>% round(2)`% were severe cases, p = `r sev_time$insects_p`).

@@

A positive history of anaphylaxis influenced the therapy of a current episode as well. Epinephrine as a first-line treatment was given less often in VIA cases when compared to other cases **if patients did not report a previous history of anaphylaxis** (`r pvaladrenprev_no`), but in patients reporting previous reactions, there was no difference in epinephrine therapy (`r pvaladrenprev`, Fig. \@ref(fig:adrenalineuse)B). 
Similarly, there were no differences in the epinephrine use between VIA and non-VIA when only severe reactions were taken into consideration (`r adren_severity_sev`). However, when we restricted the analysis to moderate anaphylaxis cases --- non-VIA patients received epinephrine more frequently than VIA (`r adren_severity_mild`).  The presence of skin symptoms during these mild reactions also was associated with a lower fraction of epinephrine treated patients (eFig. \@ref(fig:rfplot)).

Patients with VIA received corticosteroids and antihistamines significantly more frequently than patients with anaphylaxis to other elicitors. On the other hand, epinephrine, beta-2 mimetics, and oxygen were given more often to patients suffering from non-VIA  (Fig. \@ref(fig:adrenalineuse)A, eFig. \@ref(fig:rfplot)).

@@

Age is an important risk factor for severe anaphylaxis [@Stoevesandt2019]. Adult patients experienced VIA more frequently. Young patients mainly suffer from food-induced anaphylaxis [@Grabenhenrich2016]. 
Emergency room (ER) admission data indicate that the frequency of insect stings hypersensitivity reactions in children is comparable to food hypersensitivity reactions (12-15% of cases of hypersensitivity reactions admitted to the ER), but pediatric anaphylaxis is triggered significantly more often by food elicitors (56% of food hypersensitivity cases vs. 5.3% of sting cases seen in the ER) [@Braganza2005]. Senior patients, on the other hand, suffer from drug-related hypersensitivity more often than insect sting hypersensitivity [@Aurich2019]. <!-- Risk factors - Concomitant disease -->
Similarly, we observed less VIA in patients with concomitant atopic diseases (eFig. \@ref(fig:supplement)) , as these patients more often present with food anaphylaxis [@Tham2019].

@@

The European Anaphylaxis Registry was supported by the Network for Online-Registration
of Anaphylaxis NORA e. V. We thank all patients, parents, and their children for their support in providing data on the occurrence of anaphylaxis for this study. We thank the study personnel for patients counseling and data entry, and we thank the members of The European Anaphylaxis Registry in detail:

@@

```{r,include=F}
tiff("efig6.tiff", width = 7, height = 4.5,units = 'in', res = 300)
fig_basic
dev.off()
```

@@

```{r,include=F}
tiff("efig7.tiff", width = 7, height = 5,units = 'in', res = 300)
matching.result 
dev.off()
```

@@

```{r,include=F}
tiff("efig8.tiff", width = 7, height = 5,units = 'in', res = 300)
figS3
dev.off()
```

@@

```{r,include=F}
tiff("efig9.tiff", width = 7, height = 4,units = 'in', res = 300)
plot_tryptase_in_cardio_cur
dev.off()
```

@@

```{r sev, fig.cap="Severity of anaphylaxis in subgroups. The severity of patients with VIA in two age groups (left), according to elicitor type (center) and according to the responsible insect species (right)"}
figS5 <- severity_joined_brown %>%
    ggbarplot(x = "grouping",
              y = "n",
              fill = "d_severity_rm",
              position = position_fill(reverse = F),
              color = "d_severity_rm",
              palette = c("#848484","#1f1f1f"))+
    facet_grid(.~subset,scale = "free_x",space = "free_x")+
    labs(fill = "Severity")+
    theme(axis.title.x =  element_blank(),
          axis.text.x = element_text(angle=20,hjust = 1),
          legend.position = "right"
    )+
    #annotate("text",x = 1.5, y = 0.5, label = c("*"),
     #        size =12)+
    guides(color = F)
figS5
```

```{r,include=F}
tiff("efig10.tiff", width = 7, height = 4,units = 'in', res = 300)
figS5
dev.off()
```

```{r rfplot, fig.cap="Therapy of anaphylaxis. A: Patients who presented with skin symptoms and VIA less often received epinephrine than if skin symptoms were absent during the reaction. B: Variable importance in the unsupervised classification between VIA and non-VIA using Random Forest classifier.", fig.height=3.5, fig.width=7.1}
figS6 <- ggarrange(
  # adren_prev+
  #   geom_segment(aes(x = 0.5,
  #                    xend = 2.5,
  #                    y = 0.33,
  #                    yend = 0.33),
  #                color="white",
  #                linetype =2),
  skin_sympt_adrenaline,
  rf_ggplot_treatment,
  nrow = 1,
  ncol = 2,
  labels=c("A","B"),
  widths = c(1,1.3)
)
figS6
```

```{r,include=F}
tiff("efig11.tiff", width = 7., height = 3.5,units = 'in', res = 300)
figS6
dev.off()
```

@@

```{r,include=F}
tiff("efig12.tiff", width = 7, height = 4,units = 'in', res = 300)
BST_masto$plot
dev.off()
```
